### 2019-03-16 manuscript.tex
updates to notebook and text related to Archean craton pole splits

\begin{figure}
\centering
\includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}. The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

This chapter utilizes the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them with grades meant to convey the confidence that the community has in these results. While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are A and B. An A rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A B rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints.

Prior to the termination of the Trans-Hudson orogeny orogeny (before 1.8 Ga), we consider paleomagnetic poles with respect to the individual Archean cratons. Following the termination of this orogeny (after 1.8 Ga), we consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.

For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone (Bates and Halls, 1991; Evans and Halls, 2010). Evans and Halls (2010) propose an Euler rotation of (51º N, 85º W, −14º CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga. I follow this interpretation and group the poles into Superior (West) and Superior (East). These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame.

>>>>

\begin{figure}
\centering
\includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

This chapter utilizes the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them with grades meant to convey the confidence that the community has in these results. While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occuring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are included and referred to as not-rated (`NR').

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), we consider paleomagnetic poles with respect to the individual Archean cratons. There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain cratons. Following the termination of this orogeny (after 1.8 Ga), we consider poles from all the respective cratons and terranes to reflect the position of all of Laurentia.

For the Superior craton, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there as substantial Paleoproterozoic rotation of the western Superior craton relative to the eastern Superior craton across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}. \cite{Evans2010a} propose an Euler rotation of (51º N, 85º W, −14º CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga. I follow this interpretation and group the poles into Superior (West) and Superior (East). These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame.

Following Laurentia's amalgamation, poles from each part of the Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Hall201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of thick plateau developed during continent-continent collision (e.g. \cite{}). Poles are also included in the composite that come from terranes that were once part of Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction.
\begin{itemize}
\item Laurentia-Greenland.  1202 1800.0   67.5 -118.5  -13.8  1000 !Rotation of Greenland to Laurentia
\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to Darabi2004
\item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin (Maloof et al., 2006 rotation)
\end{itemize}
